---
title: Population pyramids yield accurate estimates of total fertility rates

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Mathew E. Hauer
  thanks: Data and code for reproducing this analysis are available in a public online repository. We thank A. Bronikowski, R. Lawler, and S. Alberts for their assistance with their primate data, and B. Jarosz and K. Devivo for feedback on earlier versions. Thanks y'all!
  affiliation: Department of Sociology, Florida State University
  
- name: Carl P. Schmertmann
  affiliation: Department of Economics, Florida State University

keywords:
- indirect estimation
- total fertility
- Bayesian models

abstract: The primary fertility index for a population, the total fertility rate (TFR), cannot be calculated for many areas and time periods because it requires disaggregation of births by mother’s age. Here we discuss a flexible framework for estimating TFR using inputs as minimal as a population pyramid. We develop five variants, each with increasing complexity and data requirements. To evaluate accuracy we test using more than 2,400 fertility schedules with known TFR values, across a diverse set of data sources --  including the Human Fertility Database, Demographic and Health Surveys, U.S. counties, and nonhuman species. We show that even the simplest and least accurate variant has a median error of only 0.09 births/woman over 2,400 fertility schedules, suggesting accurate TFR estimation over a wide range of demographic conditions. We anticipate that this framework will extend fertility analysis to new subpopulations, time periods, geographies, and even species. To demonstrate the framework's utility in new applications, we produce subnational estimates of African fertility levels, reconstruct historical European TFRs for periods up to 150 years before the collection of detailed birth records, and estimate TFR for the U.S. conditional on race and household income.
  
bibliography: mybibfile
output: rticles::asa_article
csl: LATEX/demography.csl
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{booktabs}
  - \usepackage{makecell}
  - \usepackage[usenames, dvipsnames]{color}
  - \usepackage{multirow}
  - \usepackage{comment}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{xcolor}
  - \newcommand{\beginsupplement}{\setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---




```{r setup, include=FALSE}

rm( list=ls() )
knitr::opts_chunk$set(echo  = TRUE)
# knitr::opts_chunk$set(cache = TRUE)


# read_chunk loads script files and
# creates labels so that they can be 
# used later in the program
#
# each script file starts with a line
# like 
#   ## @knitr libraries
# that creates a label matching the
# script's filename

library(knitr)
read_chunk('../R/SCRIPTS/load-libraries.R')
read_chunk('../R/SCRIPTS/load-constants.R')
read_chunk('../R/SCRIPTS/load-hmd-hfd-data.R')
read_chunk('../R/SCRIPTS/merge-results.R')
read_chunk('../R/SCRIPTS/load-nonhuman-data.R')
read_chunk('../R/SCRIPTS/load-us-county-data.R')
read_chunk('../R/SCRIPTS/assemble-all-error-summaries.R')
read_chunk('../R/SCRIPTS/plot-comparisons-over-all-data.R')
read_chunk('../R/SCRIPTS/plot-nonhuman-estimates.R')
read_chunk('../R/SCRIPTS/plot-multipanel-summary.R')
read_chunk('../R/SCRIPTS/plot-us-county-map.R')
read_chunk('../R/SCRIPTS/plot-errors-by-q5.R')
read_chunk('../R/SCRIPTS/plot-historical-hmd-estimates.R')
read_chunk('../R/SCRIPTS/plot-tfr-by-income.R')
read_chunk('../R/SCRIPTS/resample-CPS-income-race.R')

```

```{r load-libraries, include=FALSE}
```

```{r load-constants, include=FALSE}
```

```{r load-hmd-hfd-data, include=FALSE}
```

```{r merge-results, include=FALSE}
```

```{r load-nonhuman-data, include=FALSE}
```

```{r load-us-county-data, include=FALSE}
```

```{r assemble-all-error-summaries, include=FALSE}
```

# Introduction
Fertility is the primary engine of global population change [@gerland2014] and is central to the United Nations' Sustainable Development Goals for female education, child and maternal mortality, gender equality, and reproductive health [@abel16]. The total fertility rate (TFR) is a critical component of population change, and scientists and practitioners use it in a wide range of applications.

The conventional technique for calculating $TFR$ is straightforward, but requires data on births disaggregated by age of mother. This makes $TFR$ incalculable for: (i) countries and regions that lack detailed birth records or quality survey data, (ii) historical populations that predate vital event registration, (iii) small-area populations for which reporting agencies mask birth records for privacy reasons, and (iv) any subpopulation not identified on official birth records, such as women in a specific income decile, religion, tribe, or occupation. The need for disaggregation of births by mother's age thus limits fertility analysis -- mainly to large populations in contemporary countries with good vital registration systems or country-periods with quality survey data on fertility. 

Demographers have proposed various indirect estimation techniques to circumvent these limitations  [@bogue64;@rele67]. However, these methods often rely on variables (mean age at marriage, percent of women ever married, etc.) that may be absent from census or survey data.  Thus existing indirect methods, much like direct calculation of $TFR$, are typically limited to areas, time periods, and populations with sufficiently detailed data. In addition, relationships between fertility and social indices can differ over time and over populations, making indirect methods error-prone when applied outside of the context from which regression coefficients were derived [@tuchfeld74;@hauer13].
 
Here we discuss a flexible framework for estimating $TFR$. We derive a suite of five $TFR$ estimators that overcome the limitations above. Two of these estimators are based on previous work [@hauer13;@schmertmann2019bayesian]; three derivations are new. Our framework uses census or survey counts of population by age and sex as inputs, and it exploits demographic relationships between $TFR$ and population pyramids. Its principles are straightforward and well known. In previous research, we demonstrated that errors for this method tend to be smaller than those for other indirect methods [@hauer13] and that minor modifications using commonly-available data lead to further improvements [@schmertmann2019bayesian]. In this article we present several new variants and offer a robust evaluation of the entire framework across a wide range of demographic situations.

<!-- Here we derive a suite of five equations to estimate $TFR$ that overcome these limitations [@hauer13;@schmertmann2019bayesian]. The framework uses census or survey counts of population by age and sex as inputs, and it exploits demographic relationships between $TFR$ and population pyramids. Its principles are straightforward and well known. Recent research has demonstrated that errors for this method tend to be smaller than those for other indirect methods [@hauer13] and that minor modifications using commonly-avaiilable data lead to further improvements [@schmertmann2019bayesian].  -->

We describe the framework's derivation and evaluate the accuracy of several variants over a wide variety of databases. We test the accuracy using known $TFR$s for `r f_comma(filter(whole, Method == "iTFR")$n + nrow(combined))` fertility schedules, spanning `r max(HFD_births$Year) - min(HFD_births$Year)` years, across various scales, mortality regimes, and fertility levels.  We also offer examples of $TFR$ estimation for new types of data.  


# Methods and Materials

## Demographic Relationships between $TFR$ and age-sex distributions

Using $f(a)$ to denote the density of fertility at exact age $a$, the period total fertility rate is $TFR= \int_{\alpha}^{\beta}\,f(a)\,da$, which is usually approximated as 

\begin{equation}
TFR = \: n\cdot \sum_{a=\alpha}^{\beta-n}F_a =\: n\cdot \sum_{a=\alpha}^{\beta-n}\frac{B_a}{W_a},
\end{equation} 

\noindent where $[\alpha,\beta)$ is the reproductive age range, $W_a$ is the mid-year population of women in the $n$-year age interval $[a,a+n)$ (hereafter called \textit{age group a}), $B_a$ is the annual number of births to those women, 
and $F_a$ is their average fertility rate.
Demographers commonly use $(\alpha,\beta,n)=(15,50,5)$, in which case there are seven age groups with fertility rates $F_{15},F_{20},\ldots F_{45}$, and $TFR=5\,\cdot\,\sum F_a$.

Data for population pyramids is also reported for age groups, usually with $n=5$. Analysis of relationships between $TFR$ and the relative numbers of women and children by age group requires consideration of several demographic factors. First, not all children born during the previous $n$ years will still be alive at the time a population is enumerated. Second, not all women who gave birth over the past $n$ years will still be alive to be counted. Third, surviving women in a given $n$-year age group at the time of enumeration were only in that age group for a fraction of the past $n$ years.

These are all familiar considerations for demographers. As we demonstrate in another paper [@schmertmann2019bayesian], a slight rearrangement of standard Leslie matrix formulas \cite[e.g.~][]{wachter2014essential} 
for age groups of width $n=5$ shows that the expected number of surviving children under five, per surviving woman in age group $a$ at the end of a five-year period is

\begin{eqnarray}
\label{eq:Ca-from-components}
C_{a}\, & = & \,\,\left[\frac{L_{a-5}}{L_{a}}\,\cdot F_{a-5}\;+\; F_{a}\right]\,\,\frac{L_{0}}{2}\label{eq:Ca}\\
 & = & \,\, TFR\,\cdot\frac{L_{0}}{5}\cdot\,\frac{1}{2}\,\left(\frac{L_{a-5}}{L_{a}}\,\cdot\phi_{a-5}\;+\;\phi_{a}\right)\nonumber \\
 & = & \,\, TFR\,\cdot s\cdot\, p_{a}\nonumber
\end{eqnarray}

\noindent where $\phi_a = \tfrac{5 F_a}{TFR}$ is the fraction of lifetime fertility occurring in age group $a$ for a synthetic cohort subject to current period rates; $L_a$ is expected person-years lived in age group $a$ in a life table with a radix $l_0=1$; $s=\tfrac{L_0}{5}$ is the expected fraction still alive among children born in the past five years; $W_a$ is the observed women in age group $a$; and $W$ is the total number of women enumerated at childbearing ages $[15, 50)$.\footnote{We assume that fertility is zero outside of this range, so $F_{10}=0$ in equation (\ref{eq:Ca-from-components}) when $a=15$.}

$C_a$ is the product of three multiplicative factors: $TFR$, child survival $s$, and an age-specific term $p_a$ that represents the proportion of lifetime fertility experienced over the past five years by females in age group $a$.
The expected total number of surviving 0-4 year olds is therefore
\begin{equation}
C\,=\,\sum_{a=15}^{45}\, W_{a}\, C_{a}\,=\, W \cdot p \cdot s \cdot TFR  \label{eq:expected total C}
\end{equation}

\noindent where $p=\tfrac{1}{W}{\sum W_a p_a}$ is the population-weighted mean of $p_a$ values.
A more intuitive version of equation (\ref{eq:expected total C}) in terms of units is 
\begin{equation*}
\underbrace{\text{Children 0--4}}_{C}\,=\, 
\underbrace{\text{women}}_{W}
\cdot 
\underbrace{\tfrac{\text{births in last 5 yrs}}{\text{lifetime births}}}_{p} 
\cdot 
\underbrace{\tfrac{\text{surviving children 0--4}}{\text{births in last 5 yrs}}}_{s}
\cdot 
\underbrace{\tfrac{\text{lifetime births}}{\text{woman}}}_{TFR}  
\label{eq:expected total C in words}
\end{equation*}

\noindent We then rearrange equation (\ref{eq:expected total C}) as an expression for TFR:

\begin{equation}
TFR\,=\,\frac{1}{s}\cdot\frac{1}{p}\cdot\frac{C}{W}
\label{eq:s0TFR}
\end{equation}

## Interpretation

The most fundamental assumption behind any estimators of period $TFR$ derived from equation (\ref{eq:s0TFR}) is that the number of young children observed in an age pyramid can, after suitable correction for child mortality, serve as a proxy for recent births to the women who are counted in that same age pyramid. 
The "age pyramid" in question could come from a national census or from a regional disaggregation of national populations. But it could also come from a survey in which one can identify households in a chosen category -- for example by education of householder, geographic location, or total income. In the latter cases we count the young children and the reproductive-age women within households that belong to the chosen category. 

As long as women and their own children are counted in the same age pyramid, changes caused by internal migration (or more generally, caused by any kind of change of category) are not a major concern. The relationships in equation (\ref{eq:s0TFR}) still hold for open populations. 

Changes of category or location can affect the \textit{interpretation} of $TFR$, however. Our measures use fertility over a five-year period, and cross-sectional input data comes from the \textit{end} of that period. If changes of status are related to fertility, then indices that condition on end-of-period status may not always capture the rates that we would most like to have. For example, if single women are likely to marry quickly after having children, then an estimator based on cross-sectional data of unmarried women would accurately estimate "period fertility of women who end up unmarried" but not "period fertility of women while unmarried". Similar issues of interpretation would arise if central city residents tended to move to suburbs a short time after becoming parents. These difficulties are conceptual rather than empirical, but for stratified subpopulations it is important to understand how conditioning on end-of-period status affects interpretation of $TFR$: indices derived from equation (\ref{eq:s0TFR}) describe the \textit{recent} fertility of those in the chosen category \textit{at the time of the cross-sectional census or survey}. 


## Five $TFR$ estimators 

We derive five methods for estimating period $TFR$ from the demographic relationship in equation (\ref{eq:s0TFR}), each with different data inputs. **\autoref{methoddeployment}** shows the input data for each variant (called $iTFR$, $xTFR$, $iTFR^+$, $xTFR^+$, and $bTFR$). It is possible to use the $iTFR$ and $xTFR$ variants with age pyramid data only. If $q_5$ estimates are also available, then it is possible to use $iTFR^+$, $xTFR^+$, and $bTFR$.

**iTFR**

The simplest approximation to equation (\ref{eq:s0TFR}) assumes that child mortality is close to zero ($s\approx 1$) over the first $n$ years of life, and that women are uniformly distributed over 35 years of reproductive ages ($p\approx \tfrac{n}{\beta-\alpha}=\tfrac{5}{35}=\tfrac{1}{7}$). Following [@hauer13] we call the resulting estimator the *implied total fertility rate (iTFR)*:

\begin{equation}
\label{eq:iTFR}
iTFR  =  \frac{\beta - \alpha}{n}\cdot \frac{C}{W}  
      =  7\cdot\frac{C}{W}\nonumber 
\end{equation}

\noindent For human populations divided into five-year age groups ($(\alpha,\beta,n)=(15,50,5)$) then $iTFR=7\cdot\tfrac{C}{W}$. For other species or other age combinations, the $\tfrac{C}{W}$ multiplier may differ based on differences in $(\alpha,\beta,n)$.



\begin{table}[]
\centering
\caption{\textbf{Characteristics of alternative $TFR$ estimators.} bTFR is a probabilistic, Bayesian version using statistical distributions for unknown demographic quantities. Other estimators are deterministic.}
\label{methoddeployment}
\begin{tabular}{llccl}
                                                &                          & \multicolumn{2}{c}{Adjust for child mortality?}                                                                                        &  \\
                                                &                          & No                                                         & Yes                                                                       &  \\ \cline{3-4}
\multicolumn{1}{c}{Use age distribution detail} & \multicolumn{1}{l|}{No}  & \multicolumn{1}{c|}{iTFR} & \multicolumn{1}{c|}{iTFR\textsuperscript{+}}               &  \\ \cline{3-4}
\multicolumn{1}{c}{for women 15-49?}            & \multicolumn{1}{l|}{Yes} & \multicolumn{1}{c|}{xTFR} & \multicolumn{1}{c|}{xTFR\textsuperscript{+},   bTFR} &  \\ \cline{3-4}
                                                &                          & \multicolumn{1}{l}{}                                       & \multicolumn{1}{l}{}  
                        & \\                                      \cline{1-5}                        
\end{tabular}
\end{table}

**xTFR**

Our second estimator uses details from the population pyramid to improve the approximation of the $\frac{1}{p}$ term in equation (\ref{eq:s0TFR}). The $iTFR$ formula uses $\tfrac{1}{p}=7$, which is correct if women enumerated in the age-sex pyramid experienced a (mortality-adjusted) average of one-seventh of lifetime fertility over the previous five years. In practice this is not exactly true, because reproductive-age women may be concentrated in high- or low-fertility age groups. For example, if the age pyramid has a high concentration of women in their late 20s and early 30s, then typical age patterns of human fertility make it likely that they have just passed through five especially high-fertility ages, that $p>\tfrac{1}{7}$, and that the multiplier $\tfrac{1}{p}<7$. Conversely, a high concentration of women 40--49 in the age pyramid implies a multiplier $\tfrac{1}{p}>7$. 

Although $\tfrac{1}{p}=7$, as in $iTFR$, often leads to small errors, it is possible to improve the estimator by using additional details from the population pyramid. In previous research [@schmertmann2019bayesian] we noted that the necessary adjustments can be large for small populations like U.S. counties that have substantial variations in the age distributions of reproductive-age women. 

In order to learn about the $\tfrac{1}{p}$ multipliers, we examined `r f_comma(nrow(tmp))` fertility schedules in the \citet[HFD,~][]{HFD} for which the true $TFR$ is known. For each country $c$ and time $t$ we calculated the average $TFR$ over the five previous years, $TFR^\ast_{ct} = \tfrac{1}{5}\sum_{k=0}^4\,TFR_{c,t-k}$, and the empirical values of $TFR^\ast$ divided by observed child-woman ratios $\tfrac{C_{ct}}{W_{ct}}$. In other words, we calculated the multipliers necessary to convert $\tfrac{C}{W}\rightarrow TFR^\ast$ under the assumption of negligible child mortality. The $iTFR$ formula assumes that this multiplier equals seven. In the HFD these multipliers are within 10% of seven (6.3-7.7) in `r f_prop2percent(within10percent)` of country-years.
<!-- , and 90\% fall between 6.02 and 7.79.  -->

As in previous empirical examples [@schmertmann2019bayesian], there is a notable correlation between the age distribution of women and the multiplier. Using the proportion of women 25--34 among those who are 15--49 ($\pi_{25-34}$) as a predictor in a simple regression with HFD data produces the approximation $TFR^\ast_{ct}/({\tfrac{C_{ct}}{W_{ct}} ) } \approx `r xtfr_alpha`  `r xtfr_beta` \,\pi_{25-34}$, which we use to define the *extended TFR* or $xTFR$ estimator:

\begin{equation}
  xTFR \, = \, \left( `r xtfr_alpha` `r xtfr_beta`\, \pi_{25-34}\right) \cdot \frac{C}{W} \label{eq:xTFR}
\end{equation}

$xTFR$ adjusts for non-uniform distributions of women within reproductive ages\footnote{The coefficient values `r xtfr_alpha` and `r xtfr_beta` are appropriate when $W$ includes women $[15,50)$. Researchers could use a similar regression procedure with HMD data to produce different coefficients for other definitions of the reproductive age span.}. For any given child-woman ratio, $xTFR$ produces a lower estimate for lifetime fertility when women are more concentrated in high-fertility age groups. 

\begin{comment}
As summarized in Table \ref{xTFR-vs-iTFR}, replacing $iTFR$ with $xTFR$ yields small but meaningful improvements in predicting total fertility. 
Over HFD schedules the interquartile range of estimation errors is narrower for $xTFR$, and the mean absolute percentage error falls from a respectable 5.4\% for $iTFR$ to an even better 3.9\% for $xTFR$. 

The numbers in the table below are from [xTFR vs iTFR.R]
in the SIDE-ANALYSIS directory


\begin{table}
\centering
\caption{Distribution of $TFR$ estimation Errors over HFD Schedules}
\label{xTFR-vs-iTFR}
\begin{tabular}{rrrrr}
 Percentile
 & \multicolumn{2}{r} Arithmetic Errors 
 & \multicolumn{2}{r} Percent Errors \\
 & $iTFR$ & $xTFR$ & $iTFR$ & $xTFR$ \\ 
  \hline
Minimum & -0.77 & -0.78 & -20.6 & -21.5 \\ 
  25\%ile & -0.08 & -0.05 & -3.9 & -2.3 \\ 
  Median & 0.01 & 0.01 & 0.5 & 0.7 \\ 
  75\%ile & 0.09 & 0.06 & 5.5 & 3.6 \\ 
  Maximum & 0.56 & 0.37 & 28.1 & 15.5 \\ 
Mean Absolute Error   & 0.11 & 0.09 & 5.4 & 3.9 \\ 
   \hline
\end{tabular}
\end{table}

\end{comment}

**$\mathbf{iTFR^+}$ and $\mathbf{xTFR^+}$**

Our derivations for $iTFR$ and $xTFR$ assume no child mortality, so that the survival multiplier in Equation \ref{eq:s0TFR} is ($\frac{1}{s}\approx 1$). These simple formulations are parsimonious and generally accurate for populations with low to moderate mortality. However, when mortality levels are higher it is logical that $TFR$ would be underestimated and that errors would increase. 

```{r SAnalysis, echo=FALSE,  message = FALSE, warning = FALSE}

# calculate correlation of s=5L0/5 with 
# 1/(1-qx_const*q5) in Swedish historical data
# qx_const probably = 0.75

SWE = read.table('../R/DATA-RAW/SWE_fltper_1x1.txt', skip=2,          header=TRUE, stringsAsFactors = FALSE) 

# replace Age with numeric
SWE$Age = 0:110

radix = 100000

SWE = SWE %>%
  filter(Age < 5) %>%
  group_by(Year) %>%
  summarize(s_obs     = sum(Lx)/(5*radix),
            q5        = 1-lx[5]/radix,
            s_approx  = 1-.75*q5)

s_corr = cor(SWE$s_obs, SWE$s_approx )

s_min  = min(SWE$s_obs)
s_min_year = SWE$Year[which.min(SWE$s_obs)]

s_max  = max(SWE$s_obs)
s_max_year = SWE$Year[which.max(SWE$s_obs)]

# 
s_error = (1/SWE$s_approx) / (1/SWE$s_obs)

```

Our third and fourth estimators, denoted $iTFR^+$ and $xTFR^+$, approximate the $\tfrac{1}{s}$ component in equation (\ref{eq:s0TFR}) using estimated under-five mortality. 
Specifically, we use $\tfrac{1}{s}\approx \tfrac{1}{1-`r qx_const`\,q_5}$, which is based on both logical and empirical relationships between life table variables. In `r nrow(SWE)` years of Swedish HMD data, with under-five survival rates ranging from `r f_num(s_min,3, p ="0")` (in `r s_min_year`) to `r f_num(s_max,3, p ="0")` (in `r s_max_year`), the approximation $\hat{s}=1-`r qx_const`\,q_5$ had a correlation `r  f_num(s_corr, 3, p ="0")` with $s$. Over these `r nrow(SWE)` years replacing true multipliers $\tfrac{1}{s}$ with approximations $\tfrac{1}{\hat{s}}$ in Equation (\ref{eq:s0TFR}) would produce very slight underestimates of $TFR$, ranging  from `r round( min(100 * (s_error-1)),1)`\% to `r round( max(100 * (s_error-1)),2)`\% with a median underestimate of `r round(quantile(100*(s_error-1), 0.5),2)`\%.

We call the variants that include simple adjustments for child mortality $iTFR^+$ and $xTFR^+$. Specifically,

\begin{eqnarray}
iTFR^+ \, & = & \left( \frac{7}{1-`r qx_const`\,q_5} \right) \cdot \frac{C}{W} \\
xTFR^+\, & = & \left( \frac{`r xtfr_alpha` `r xtfr_beta` \pi_{25-34}}{1-`r qx_const`\,q_5} \right) \cdot  \frac{C}{W}
\end{eqnarray}


**A probabilistic model: $\mathbf{bTFR}$**

Our fifth estimator, $bTFR$, is a fully probabilistic Bayesian model. It uses details from the population pyramid about female age structure within reproductive ages, and it requires an estimate of under-five mortality. 

The Bayesian approach treats the number of children as a Poisson random variable, and treats the demographic quantities in Equations (\ref{eq:Ca-from-components}) and (\ref{eq:expected total C}) as uncertain. We derive prior distributions for the fertility and mortality patterns that determine $p$ and $s$ from large demographic databases [@HFD;@HMD]. We define the $bTFR$ estimator as the median of the marginal posterior distribution of a population's $TFR$, conditional on observed $C$ and $(W_{15}\ldots W_{45})$. Because they are probabilistic, $bTFR$ estimators automatically produce uncertainty measures as well as point estimates. We describe this method in detail in another paper [@schmertmann2019bayesian]. We briefly summarize here.

*Fertility Parameters*

The proportion of lifetime fertility experienced by women in the five years before a census or survey depends on their age distribution (which is observed), and on the relative levels of fertility in different age groups (which are uncertain).
In order to model this uncertainty we decompose the fertility schedule for 5-year age groups into level and shape components, $(F_{15},...,F_{45})= \frac{TFR}{5} \cdot (\phi_{15},...,\phi_{45})$. 
As in [@schmertmann2019bayesian] we model the proportions $\phi_{15}...\phi_{45}$ in terms of log odds,  $\gamma_{a}=\ln(\frac{\phi_a}{\phi_{15}})$ for $a=15...45$ and then translate as $\phi_a(\gamma)= \frac{\exp(\gamma_{a})}{\sum_{z}\exp(\gamma_{z})}$. By construction, these seven $\phi_a$ values are positive and sum to one.

Our model for the $\gamma$ indices is $\gamma=m+X\beta$ where $m=(\begin{smallmatrix} 0 &1.39 &1.59 &1.23 &0.45 &-0.89 &-3.44\end{smallmatrix})^\prime$ and $X= \left(\begin{smallmatrix}0 & 0.27 & 0.54 & 0.73 & 0.88 & 1.04 & 1.52\\0 & 0.32 & 0.51 & 0.51 & 0.35 & 0.05 & -0.72 \end{smallmatrix}\right)^{\prime}$ are constants derived from empirical data and $\beta$ are unknown shape parameters.\footnote{The highest values in $m$ correspond to age groups $[20,25)$ and $[25,30)$, so that those age groups represent the highest shares of lifetime fertility. Because the first column of $X$ has monotonically increasing values, the first element of $\beta$ affects the mean age of childbearing: higher values raise the log odds of late relative to early fertility. Similarly, the second element of $\beta$ affects the variance of age-specific fertility, with higher $\beta_2$ causing higher concentration of fertility in the 20s and lower variance.}  See [@schmertmann2019bayesian] for details. When combined with a prior distribution $\beta\sim N(0,I_2)$ this model assigns higher prior probabilities to fertility age patterns that are more similar to those in the HFD and the U.S. Census Bureau's International Database [@CensusIDB]. 

We use an uninformative prior for $TFR$: $TFR\sim \text{Uniform}(0,20)$. This allows the level of total fertility to be determined almost completely by the data, rather than by prior assumptions.


*Mortality Parameters*

We model child and adult mortality with a relational mortality model [@wilmoth2012flexible], in which two parameters describe the complete pattern of mortality rates by age: the probability of death before age five ($q_5$), and a shape parameter $k$ with typical values between -2 and +2. The model uses fixed constants $\left\{a_x,b_x,c_x,v_x\right\}$ derived from mortality schedules in the Human Mortality Database:

\begin{equation}
\ln\,\mu_{x}(q_{5},k) = a_{x}+b_{x}\,\left[\ln\, q_{5}\right]+c_{x}\,\left[\ln\, q_{5}\right]^{2}+v_{x}\, k\quad,\quad x=0,1,5,10 \ldots 45
\end{equation}

\noindent We use standard demographic calculations to convert these log mortality rates into the $L_{a}$ values in Equation (\ref{eq:Ca}).

To account for possible errors in estimated mortality, we use a beta distribution for the true value of $q_5$. Our prior is $q_5 \sim Beta(a,b)$
with $a$ and $b$ such that $P[q_{5}<\tfrac{1}{2}\,\min(\hat{q}_{5})]\:=\: P[q_{5}>2\,\max(\hat{q}_{5})]\:=\:.05$. This prior allows for a considerable amount of possible error in the $q_5$ estimate used as input: it assigns a 90% prior probability that the true value of $q_5$ is between one-half and twice its estimated value. For the shape parameter $k$ our prior is $k \sim N(0,1)$. We assume *a priori* that mortality parameters $q_5$ and $k$ are independent.

*Complete bTFR Model*

Any set of parameters ($TFR$, $\beta$, $q_5$, $k$) implies specific values $C_a$ in equation (\ref{eq:Ca}). The expected number of surviving children to the $W_{a}$
women observed in age group $a$ is $W_{a}C_{a}$, and the observed
number of surviving children to these women is a Poisson random variable with mean 
$W_{a}C_{a}.$\footnote{A Poisson model assumes equality between the mean and variance of the number of surviving children. This strong assumption generally does not hold in practice \cite[e.g.~][Figure 1]{Barakat_36_26}. Individual heterogeneity in age-specific rates would tend to produce overdispersion (variance $>$ mean), while strong social norms about childbearing might tend to produce underdispersion (variance $<$ mean). In a comprehensive empirical study of cohort parity, Barakat \citeyear{bilal_barakat_revisiting_2014} provides evidence for effects in both directions. Underdispersion is more common at low parities, and vice versa. The good performance of the $bTFR$ estimator in our empirical tests suggests that the Poisson model is adequate for estimating $TFR$ from age-sex distributions.} Assuming statistical independence across maternal age groups, the total number of children 
also has a Poisson distribution: 
\begin{equation}
C|W,TFR,\beta,q_5,k \sim Poisson \Bigg[ \sum_{a=15}^{45}W_a\, C_a(TFR,\beta,q_5,k) \Bigg] \label{eq:poisson distribution of C}
\end{equation}

The posterior distribution of parameters conditional on age pyramid data is therefore
\begin{equation}
P(TFR,\beta, q_5,k | C,W) \propto L(C|W,TFR,\beta,q_5,k) f_\beta(\beta)f_q(q_5)f_k(k)  \label{eq:joint posterior}
\end{equation}

\noindent where the likelihood $L$ is Poisson and the $f$ functions represent
the prior densities for unknown parameters. The flat prior for $TFR$ does not affect the posterior distribution over the range $TFR\in[0,20]$.

The marginal posterior for TFR provides the relative probabilities
of alternative fertility levels, given the number of children $C$
and the counts of women $W_{15}\ldots W_{45}$ in the observed age pyramid. We sample from the joint posterior distribution (\ref{eq:joint posterior}) using Markov Chain Monte Carlo (MCMC) methods, programmed in \textit{Stan} [@stanjss2017] using the \textit{rstan} package in \textit{R} [@rstanpackage;@Rlanguage], and we estimate the marginal posterior of $TFR$ using the empirical density of sampled $TFR$ values.


# Data


In order to evaluate the accuracy of the five estimators over as many different data situations as possible, we use four data sources that together comprise `r f_comma(filter(whole, Method == "iTFR")$n + nrow(combined))` fertility schedules.

## Human Fertility Database/Human Mortality Database
```{r , echo=FALSE, warning=FALSE, message = FALSE, include = FALSE}
# options(knitr.kable.NA = '')
hfdcountries <- tribble(
  ~Code, ~Country,
  "AUT", "Austria",
  "BGR", "Bulgaria",
  "BLR", "Belarus",
  "CAN", "Canada",
  "CHL", "Chile",
  "HRV", "Croatia",
  "CZE", "Czechia",
  "DEN", "Denmark",
  "EST", "Estonia",
  "FIN", "Finland",
  "FRATNP", "France",
  "DEUTNP", "Germany",
  "HUN", "Hungary",
  "ISL", "Iceland",
  "ITA", "Italy",
  "JPN", "Japan",
  "LTU", "Lithuania",
  "NLD", "Netherlands",
  "NOR", "Norway",
  "POL", "Poland",
  "PRT", "Portugal",
  "RUS", "Russia",
  "SVK", "Slovakia",
  "SVN", "Slovenia",
  "ESP", "Spain",
  "SWE", "Sweden",
  "CHE", "Switzerland",
  "TWN", "Taiwan",
  "UKR", "Ukraine",
  "GBRTENW", "England and Wales",
  "GBR_NIR", "Northern Ireland",
  "GBR_SCO", "Scotland",
  "USA", "USA"
)


hmdtab <- HFD_tfr  %>%
  group_by(Code) %>%
  dplyr::summarise(Start = min(Year),
                   End = max(Year),
                   yearlength = n()) %>%
  left_join(., hfdcountries)
hmdtab2 <- hmdtab %>%
  mutate(`Data Avail. (# of Years)` = paste0(Start, "-", End, " (", yearlength,")")) %>%
  dplyr::select(Country,`Data Avail. (# of Years)` )

hmdtab2[nrow(hmdtab2)+1,] <- ""

dd2 <- cbind(hmdtab2[1:16, ], hmdtab2[17:32, ])

```
We first benchmark the estimators against the \citet{HFD}, the most complete and accurate dataset available on current and historical patterns of human fertility. The HFD covers fertility schedules for `r length(unique(HFD_births$Code))` countries between `r min(HFD_births$Year)` and `r max(HFD_births$Year)`, containing `r f_comma(sum(hmdtab$yearlength))` country-years of age-specific and total fertility rates. These are listed in **Supplementary Table 1**.


For each country-year in the HFD, we link the corresponding population data (i.e., the age pyramid), and the $q_5$ value from the \citet{HMD}. When joined, the HFD/HMD data includes true target values for $TFR$ and input data ($C, W_{15}\ldots W_{45}, q_5$) for the five estimators. 


## Demographic and Health Surveys

HFD/HMD data is highly accurate, but it covers a fairly narrow range of demographic conditions. HFD populations are mainly in contemporary developed countries with relatively low fertility and mortality rates. 

To evaluate the estimators under a broader set of conditions, we also test against Demographic and Health Survey (DHS) data. DHS data includes nationally representative household surveys for monitoring population health in `r length(unique(filter(big, source == "DHS")$Code))` countries and will produce TFR estimates based on the sample. For `r nrow(filter(big, source == "DHS"))` country-periods we use the DHS API [@DHSAPIcite] to download estimated $TFR$, number of women aged 15-49, number of children under age 5, and $q_5$. We use IPUMS-DHS [@IPUMSDHS] to download the corresponding numbers of women by five-year age group. The combined DHS data includes a (noisy but unbiased) estimate of the target value for $TFR$, and the ($C, W_{15}\ldots W_{45}, q_5$) input data for the five estimators. 

## U.S. County Fertility

Combining HFD/HMD and DHS data allows us to evaluate our methods at the national level. 
To evaluate the estimators at sub-national geographies, we use age-sex distributions from the 2010 Decennial Census to produce $xTFR$ and $iTFR$ estimates for every county in the U.S.. We evaluate the accuracy of these estimates by comparing to published county-level $TFR$ values from the National Center for Health Statistics (NCHS), obtained via the Center for Disease Control (CDC)'s Wide Ranging Online Data for Epidemiological Research (WONDER) tool.

The National Center for Health Statistics (NCHS) publishes highly accurate, sub-national fertility information in the United States. However, for privacy reasons the National Center for Health Statistics does not publish fertility information for U.S. Counties with populations less than 100,000. As a result, we can compare estimates to true county-level $TFR$ values for only `r countyeval$tot` of the approximately 3000 U.S. Counties.

## Nonhuman data

The basic $iTFR$ variant only requires information on a population's age-sex structure. Population pyramids are not limited to humans, so $iTFR$ could in principle be used to estimate total fertility in nonhuman populations if age-sex distributions were accurate. To test this proposition we assemble age-specific fertility and population data for eleven nonhuman species (eight wild and one captive primate species, one wild lion species, and one wild seal species) reported in **\autoref{nonhuman-datasources}**. 

These populations vary substantially, allowing us to assess whether the method works across species with very different levels and age patterns of fertility. In contrast to humans, for whom scientists typically demarcate menarche and menopause at ages 15 and 50 years, respectively, menarche among the eleven species ranges from a low of age `r min(combined$fertstart)` for Sifakas (*P. verreauxi*) and African lions (*P. leo*) to a high of `r max(combined$fertstart)` years for Chimpanzees (*P. troglodytes*). Reproductive age spans range from a low of `r min(combined$fertlength)` years for Thomas Langurs (*P. thomasi*) to a high of `r max(combined$fertlength)` years for Chimpanzees. These species display reproductive age spans, fertility schedules, and TFRs that differ greatly from humans. 

Demographic data for nonhuman
populations is collected very differently from human data. Living individuals must be directly observed, their ages must be estimated, and their sex may or may not be known. 
Despite these differences, it is interesting to evaluate the effectiveness of estimating lifetime fertility from age structure for other species. 

\begingroup
\renewcommand{\arraystretch}{0.75}
\begin{table}[]
\caption{\textbf{nonhuman fertility data.} $\alpha$ and $\beta$ are the first and last ages with observed non-zero fertility rates for each species. $_{1}P_0$ is the number of individuals less than one year old. $_{\beta-\alpha}W_{\alpha}$ is the number of females of reproductive age. $TFR$ is observed total fertility. Sources are 
(a) \citet{bronikowski16,bronikowskidata}, 
(b) \citet{wich2007demography},
(c) \citet{ha2000demographic},
(d) \citet{barlow1991modeling},
(e) \citet{packer1998reproductive}.}
\label{nonhuman-datasources}
\begin{tabular}{lrrrrrr}
\cline{1-7}
Species and Source        & Location   & $\alpha$ & $\beta$ & $_1P_0$ & $_{\beta-\alpha}W_\alpha$ & TFR  \\ \cline{1-7}
Sifaka (\textit{P. verreauxi})\textsuperscript{a}          & Madagascar & 2        & 31      & 708     & 2073                      & 10.67                  \\
African Lion (\textit{P. leo})\textsuperscript{e}          & Tanzania   & 2        & 17      & 2643    & 3819                      & 10.38          \\
Capuchin (\textit{C. capucinus})\textsuperscript{a}        & Costa Rica & 5        & 24      & 189     & 423                       & 9.82                   \\
Baboon (\textit{P. cynocephalus})\textsuperscript{a}       & Kenya      & 4        & 23      & 1098    & 2465                      & 9.37                   \\
Muriqui (\textit{B. hypoxanthus})\textsuperscript{a}       & Brazil     & 7        & 40      & 376     & 1347                      & 8.89                   \\
Blue Monkey (\textit{C. mitis})\textsuperscript{a}         & Kenya      & 4        & 30      & 486     & 1542                      & 8.69                   \\
Gorilla (\textit{G. beringei})\textsuperscript{a}          & Rwanda     & 8        & 40      & 258     & 1049                      & 8.00                   \\
Macaque (\textit{M. nemestrina})\textsuperscript{c}        & captive    & 3        & 22      & 2368    & 7494                      & 6.91                \\
Chimpanzee (\textit{P. troglodytes})\textsuperscript{a}    & Tanzania   & 11       & 49      & 189     & 1195                      & 6.29                   \\
Northern Fur Seal (\textit{C. ursinus})\textsuperscript{d} & --       & 3        & 19      & 2096    & 6436                      & 6.10            \\
Thomas Langur (\textit{P. thomasi})\textsuperscript{b}      & Indonesia  & 4        &   11      & 161     & 305                       & 3.67  \\
\cline{1-7}
\end{tabular}\label{nonhumantable}
\end{table}
\endgroup

# Evaluation

## Overall Error

We begin by investigating $TFR$ estimation errors over all available test cases for human data. 
**\autoref{comparison}** reports the 10th, 50th, and 90th  percentile of the absolute error ($\left| est-obs \right|$) for all methods in the HMD/HFD, DHS, and U.S. County datasets. 

All methods produce quite accurate results using all three data sources, for almost all of the populations under study. Only the $xTFR$ variant for DHS data produces estimates with the median absolute error above 0.25 births/woman. Median absolute errors are less than 0.10 births/woman in most datasets.

In general, more sophisticated estimators have lower average errors. Ordering by median errors shows that the simplest estimator, $iTFR$, has the poorest (although still very good) performance, that adding age structure detail and/or mortality corrections ($xTFR$, $iTFR^+$, $xTFR^+$) reduces errors further, and that most complex model ($bTFR$) has the smallest average errors. 

There is one important exception to the general "fancier is better" rule. In populations with high mortality and fertility rates -- exemplified in our data by DHS countries -- adding details about the age distribution of reproductive-age women while ignoring child mortality (i.e., changing $iTFR \rightarrow xTFR$) actually makes estimates worse. Why? Recall from Equation (\ref{eq:s0TFR}) that the true  $\tfrac{C}{W}\rightarrow TFR$ multiplier has two parts: $\tfrac{1}{s}$ for child mortality correction, and $\tfrac{1}{p}$ for age structure correction. $iTFR$ assumes $\tfrac{1}{s}=1$ and $\tfrac{1}{p}=7$. In a high-mortality, high-fertility population the true values of the two components actually tend to be greater than 1 (more than one birth per living child) and less than 7 (children born recently represent less than 1/7 of the population's lifetime births, because there are relatively few old women). In such situations $iTFR$ underestimates one component of the multipler ($\tfrac{1}{s}$) and overestimates the other ($\tfrac{1}{p}$), while $xTFR$ underestimates ($\tfrac{1}{s}$) and gets ($\tfrac{1}{p}$) about right. Compensating errors for $iTFR$ mean that it actually performs better than the more sophisticated $xTFR$ approach in a high-mortality, high-fertility setting. 

This leads to one important caveat. If mortality estimates are unavailable, then $iTFR$ is preferable to $xTFR$ in a population with suspected high mortality. 


```{r plot-comparisons-over-all-data, echo=FALSE, message = FALSE, warning = FALSE, dev = c("png"),dpi=300,fig.cap= paste("**Absolute Error (Births/Woman) in TFR over alternative methods and datasets.** Solid dots are at median Error for each method and dataset. Horizontal bars extend from the 10th-90th percentile of Error. Numbers in parentheses indicate the count of schedules for which it is possible to use each method. Median errors are within 1/10th of a birth across all data sets. \\label{comparison}")}

```

```{r plot-nonhuman-estimates, echo=FALSE, message = FALSE, warning = FALSE, out.width = "60%", dev = c("png"),dpi=300,fig.align="center", fig.cap= paste("\\textbf{Fertility estimates for animal populations}. $iTFR$ estimates from nonhuman age-sex distributions. Observed TFR versus $iTFR$ estimates for the species listed in \\autoref{nonhumantable}. Estimates match observations along the 45-degree line; dashed lines represent +/- 10\\% errors.\\label{primates}")}
```


## Errors for Nonhuman Species

To test the generalizability of the method, we examine the accuracy of the $iTFR$ estimator in `r nrow(combined)` nonhuman populations described in (**\autoref{nonhumantable}**). We find that $iTFR$ accurately estimates total fertility among these species (**\autoref{primates}**). These results suggest the method captures fundamental properties that govern mammalian fertility, and that it could be valuable in other studies with nonhuman populations. 


## Sensitivity to TFR Level, Population Size, and Time Period

```{r plot-multipanel-summary, echo=FALSE, message = FALSE, warning = FALSE, fig.height=6.5,dev = c("png"),dpi=300, fig.cap=paste("**Estimated TFR from Population Pyramids.** Performance of three variants in HFD and DHS data. (a, d, g) use $iTFR$; (b, e, h) use $xTFR$; (c, f, i) use $bTFR$; (a, b, c)  plot estimated TFR against the observed 5-year average TFR. The solid line is $Y=X$, and the dashed lines are $\\pm 10\\%$. (d, e, f) illustrate percent error against population size. The dashed lines represent errors of $\\pm 10\\%$. (g, h, i) plot percent errors against the year in which the population pyramid is observed. (j) plots the distribution of algebraic errors for each method ($est-obs$). (k) plots the distribution of absolute algebraic errors. (l) plots the distribution of absolute percent errors. For all variants, estimates are accurate over many scales and times. \\label{multipanel-summary}")} 
```

**\autoref{multipanel-summary}** reports the accuracy of $iTFR$, $xTFR$, and $bTFR$ estimators for the HFD/HMD and DHS data combined for different levels of TFR, different population sizes, and different historical periods. **\autoref{error-table}** reports summary measures of the accuracy for all five estimators using the same combined HFD/HMD and DHS data as well as U.S. counties. Overall, we find good agreement between estimated and observed TFRs for all five estimators (**\autoref{multipanel-summary}**, a, b, c and **\autoref{error-table}**.

Demographic estimators are typically more accurate for larger populations (due to the law of large numbers) and for more recent time periods (due to improved data collection practices). However, we find that error rates are independent of population size  (**\autoref{multipanel-summary}**, d, e, f) and are fairly stable across time (**\autoref{multipanel-summary}**, g, h, i), suggesting scale and temporal independence uncommon in other indirect methods.

Even the simplest and least accurate of the five variants, $iTFR$, predicts the total fertility rate with absolute errors of less than `r prettyNum(round(itfrabs50,2), digits=2)` births/woman in half of the HFD and DHS populations, and less than `r round(itfrabs90,2)` births/woman in 90\% of the populations (**\autoref{error-table}**). Absolute percent errors for $iTFR$ are also quite small relative to most indirect demographic estimators: 50\% of errors are within $`r sprintf("%.1f", itfrape50)`$\% of the true TFR, and 90\% are within $`r sprintf("%.1f", itfrape90)`$\%. As shown in **\autoref{multipanel-summary}** and **\autoref{error-table}**, the additional information contained in the $xTFR$ and $bTFR$ estimators produce even smaller errors. In short, we find that for national populations in countries with accurate data and (mostly) low mortality, accurate $TFR$ estimates from population pyramids are possible.

```{r, echo=FALSE, warning=FALSE, message = FALSE, results='asis'}

# comes from 'assemble-all-error-summaries.R`

kable(allerrortable, "latex", booktabs = T, caption = "\\label{error-table}\\textbf{Summary statistics for the five variants using data from the HMD/HFD and DHS and U.S. counties.} APE is the Absolute Percent Error.") %>%
  kable_styling(latex_options = c("scale_down")) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") 
```



## Errors in Subnational Estimates

It is possible that coefficients derived at the National level could prove ineffective in estimating fertility at the subnational level (eg. @tuchfeld74). **\autoref{error-table}** reports the errors associated with both the $iTFR$ and $xTFR$ methods for U.S. counties that also have corresponding observed total fertility rates. These errors are on par with the errors observed using the HMD/HFD data, demonstrating important consistency in low error rates across a variety of scales.

## Sensitivity to Child Mortality

```{r plot-errors-by-q5, echo=FALSE, message = FALSE, warning = FALSE, dev = c("png"),dpi=300,fig.height = 5.5, fig.cap=paste("**Absolute Percent Errors against $q_5$ values in the HMD and DHS.** We compare the performance of the five variants against observed $q_5$ mortality rates. As $q_5$ values increase, APE also increases in the $iTFR$ and $xTFR$ variants. This is corrected in the $iTFR^+$, $xTFR^+$, and $bTFR$ variants, which incorporate estimated child mortality. \\label{errors-by-q5}")}
```

Our derivations for the $iTFR$ and $xTFR$ assume that child mortality is negligible, and those estimators are likely to underestimate fertility when it is not. 
We show this relationship in **\autoref{errors-by-q5}** using data from the HMD/HFD and DHS, where $iTFR$ and $xTFR$ produce increasing errors as child mortality increases. Notice that the $iTFR^+$, $xTFR^+$, and $bTFR$ variants largely correct for higher mortality levels. Thus, in countries or periods with high infant and child mortality, the $iTFR^+$, $xTFR^+$, and $bTFR$ variants are more appropriate.



## Extensions {#sec:extensions}

Accurate estimation of $TFR$ from age-sex pyramids greatly expands our ability to estimate fertility across varying geographies, time periods, and subpopulations. To demonstrate the flexibility of the method, we produce TFR estimates for three cases in which direct $TFR$ estimation is impossible. 

*Subnational Fertility in Africa*

We use data from WorldPop's gridded population age-structure data for the year 2015 [@tatem2013millennium] in conjunction with $q_5$ estimates from the United Nations World Population Prospects 2017 [@united2017world] to produce subnational estimates of $iTFR^+$ for Africa. Scholars have increasingly published subnational estimates of demographic indicators to monitor progress towards the UN's Sustainable Development Goals (SDGs) [@osgood2018mapping;@golding2017mapping;@graetz2018mapping] and our framework allows a nearly unprecedented level of geographic detail regarding fertility (**\autoref{WorldPop}**). 

```{r plot-africa-map, echo=FALSE,  message = FALSE, warning = FALSE,dev = c("png"),dpi=300, fig.cap= paste("**Estimating subnational fertility rates.**  We use the $iTFR^+$ method to estimate subnational total fertility rates for the African continent for 2010-2015 using data from WorldPop and the United Nations WPP 2017.  \\label{WorldPop}")}

## the map is generated by make-africa-map.R,
## run before markdown because it takes a
## very long time to render the map

iTFRafrica <- pdf_render_page('../MANUSCRIPT/FIGURES/top_africa4.pdf', page = 1, dpi = 600, numeric = FALSE)

ggdraw() + draw_image(iTFRafrica)
```

*Historical Fertility in Europe*

We also extend our analysis of the Human Mortality Database by producing fertility estimates for all `r nrow(read.csv('../R/DATA-PROCESSED/HMDHFD-allbTFR_06172017.csv'))` country-years of data in the HMD (an additional 1000 country-years' of estimates prior to the collection of detailed birth records). From this large historic volume of data, we highlight our findings in four example countries: France, Italy, the Netherlands, and Sweden (**\autoref{Historic}**). Sweden began tabulating the detailed birth records necessary for $TFR$ calculation in 1891, France in 1946, the Netherlands 1950, and Italy in 1954. However, these countries collected both mortality and age-sex data considerably earlier (1751 for Sweden, 1816 for France, 1850 for the Netherlands, and 1872 for Italy). By using the $bTFR$ method, we can reconstruct historical TFRs to create a time series of fertility data well before age-specific birth collection began, significantly expanding our ability to explore historical fertility patterns from up to 250 years ago.

```{r plot-historical-hmd-estimates, echo=FALSE,  message = FALSE, warning = FALSE, dev = c("png"),dpi=300,fig.cap= paste("**Estimating historical fertility.**  $bTFR$ estimates of period fertility rates in four European countries using  HMD historical age-sex and child mortality data. Shaded regions represent 90 percent posterior probability intervals; open circles are observed TFRs from the HFD. Red stars are average TFRs over the preceding five-year period for the earliest series with a vital records source from the Human Fertility Collection \\citep{HFC}.\\label{Historic}")}

```

*Fertility by Income and Race in the United States*

The complex connections between fertility and household income have always interested demographers [@becker1960economic;@jones2008fertility]. However, analysis is limited because birth certificates do not include economic information, and very few surveys include both income and fertility questions. Estimation of fertility levels from population composition can expand our ability to learn about fertility-income correlations in different social groups.  

(**\autoref{Income}**) provides an example. We use the $xTFR$ method to estimate TFRs conditional on both race (White/Non-White) and household income level in the United States, using data on the age-sex composition of households in different income strata from the Current Population Survey (CPS). Data for this figure comes from aggregated 2010--2018 March Economic Supplements, downloaded from IPUMS [@IPUMSCPS]. Indirect methods based on age-sex data allow us to produce estimates of $TFR$ by income groups, and to further disaggregate by race. Among both Whites and Non-Whites, there is a definite U-shaped relationship, with the highest fertility levels in the poorest and richest American households. 

Here we provide these examples only as a proof of concept: indirect fertility estimation identifies intriguing relationships that would not be estimable by other means, and which clearly merit further study. 


```{r include=FALSE}
## the plot is generated in advance 
## by make-CPS-income-race-plot.R

# CPS <- pdf_render_page('FIGURES/CPS-income-race.pdf', page = 1, dpi = 600, numeric = FALSE)
# 
# ggdraw() + draw_image(CPS)
# NOTE: To load data, you must download both the extract's data and the DDI
# and also set the working directory to the folder with these files (or change the path below).

# library(tidyverse)
# 
# if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
# 
# ddi <- read_ipums_ddi("DATA-RAW/CPS/cps_00003.xml")
# 
# # select March supplement data only
# 
# data <- read_ipums_micro(ddi) %>%
#           filter(  ASECFLAG==1 ) 

```

```{r resample-CPS-income-race, echo=FALSE,  message = FALSE, warning = FALSE, dev = c("png"),dpi=300,fig.cap= paste("**Total Fertility ($xTFR$) by race and household income level.** We estimate TFR from age-sex distributions within race-income categories using the U.S. Current Population Survey (CPS) March Social and Economic Supplements, combined over 2010-2018. White corresponds to those reporting their race as White only; Non-White corresponds to all other survey respondents. Shaded regions represent confidence intervals (5%-95%) estimated from 1000 bootstrap samples in which households in each rae-income category are drawn randomly with replacement from the set of all CPS households. \\label{Income}")}

## the plot is generated in advance 
## by make-CPS-income-race-plot.R

# CPS <- pdf_render_page('FIGURES/CPS-income-race.pdf', page = 1, dpi = 600, numeric = FALSE)
# 
# ggdraw() + draw_image(CPS)
# NOTE: To load data, you must download both the extract's data and the DDI
# and also set the working directory to the folder with these files (or change the path below).

# library(tidyverse)
# 
# if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
# 
# ddi <- read_ipums_ddi("DATA-RAW/CPS/cps_00003.xml")
# 
# # select March supplement data only
# 
# data <- read_ipums_micro(ddi) %>%
#           filter(  ASECFLAG==1 ) 
# 
# write_csv(data, "DATA-PROCESSED/IPUMS_CPS.csv")
# 
# data<- read_csv("DATA-PROCESSED/IPUMS_CPS.csv")

# inc_cutoffs = c(-Inf, 25, 50, 75, 100, 150, 200, 250, 300, Inf)
# inc_labels = c('<25','25-50','50-75','75-100',
#                '100-150','150-200',
#                '200-250','250-300','>300')
# 
# mydata = data %>%
#           mutate(income_cat = cut(HHINCOME/1000, 
#                                   breaks=inc_cutoffs,
#                                   labels=inc_labels),
#                  age_cat    = 5*floor(AGE/5),
#                  race_eth   = case_when(
#                                 RACE == 100 ~ "White",
#                                 TRUE        ~ "Non-White"
#                              )) %>%
#           group_by(sex=SEX,race_eth,age=age_cat,income=income_cat) %>%
#           summarize(n=sum(ASECWT), unwtd=n()) %>%
#           group_by(race_eth,income) %>%
#             summarize( C   = sum(n[age==0]),
#                        unwtd_W = sum(unwtd[(age %in% 15:45) & (sex==2)]),
#                        W15 = sum(n[(age==15) & sex==2]),
#                        W20 = sum(n[(age==20) & sex==2]),
#                        W25 = sum(n[(age==25) & sex==2]),
#                        W30 = sum(n[(age==30) & sex==2]),
#                        W35 = sum(n[(age==35) & sex==2]),
#                        W40 = sum(n[(age==40) & sex==2]),
#                        W45 = sum(n[(age==45) & sex==2]),
#                        W = W15+W20+W25+W30+W35+W40+W45,
#                        pi2534 = (W25+W30)/W,
#                        CWR = C/W,
#                        iTFR = 7*CWR,
#                        xTFR = (10.65 - 12.55*pi2534) * CWR) %>%
#                  dplyr::select(race_eth,income,unwtd_W,C,W:xTFR)
#               
# ## manually construct smoothed lines, just to make them
# ## look a little nicer over chunky intervals
# tmp = mydata %>% 
#         dplyr::select(income, xTFR) %>%
#         mutate(income = as.numeric(income))
# 
# income_grid = seq(1,9,.10)
# 
# Wfit = loess(xTFR ~ income, data=tmp, span=1,
#              subset= (race_eth=='White')) %>%
#        predict(newdata=expand.grid(income=income_grid)) %>%
#        as.numeric()
# 
# NWfit = loess(xTFR ~ income, data=tmp, span=1, 
#              subset= (race_eth=='Non-White')) %>%
#   predict(newdata=expand.grid(income=income_grid)) %>%
#   as.numeric()
# 
# mysmooth = expand.grid( income = income_grid,
#                         race_eth = c('White','Non-White'))
# 
# mysmooth$xTFR  = c(Wfit, NWfit)
# 
# ggplot(data=mydata, 
#        aes(x=income,y=xTFR,shape=race_eth, group=race_eth,color=race_eth)) +
#     geom_line(linetype=2) +
#     geom_point(size=3.5) +
#     scale_shape_manual(values=c('N','W'), guide=FALSE) + 
#     scale_y_continuous(breaks=seq(1.6,2.6,0.2)) +
#   geom_smooth(span = 1, aes(fill = race_eth, color = race_eth, group = race_eth, shape = race_eth)) +
#     # geom_line(data=mysmooth, lwd=2) +
#     geom_hline(yintercept = 2) +
#     theme_bw() +
#     # theme(axis.text = element_text(face='bold', size=14),
#     #       axis.title=element_text(size=14,face="bold")) +
#     labs(x='Household Income in Thousands',
#          y='TFR') +
#     scale_color_manual(values=c('red','blue'), guide=FALSE) +
#     geom_text(x=7.5,y=2.4,label='Non-White', color='black') +
#     geom_text(x=8.0,y=1.8,label='White', color='black') +
#   guides(fill=FALSE)

ggplot(data=case, 
       aes(x=this_income,y=xTFR,
           shape=this_race, group=this_race, color=this_race)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95, fill = this_race), color = NA, alpha = 0.3) +
  geom_line(linetype=2) +
  geom_point(size=3.5) +
  # geom_segment(data=case, aes(x=this_income, y=Q05, xend=this_income, yend=Q95),
  #              lwd=1.5, alpha=.50) +
  scale_shape_manual(values=c('W','N'), guide=FALSE) + 
  scale_y_continuous(breaks=seq(1.6,2.6,0.2)) +
  #  geom_line(data=mysmooth, lwd=2) +
  geom_hline(yintercept = 2) +
  theme_bw() +
  theme(
  # axis.text = element_text(face='bold', size=14),
        axis.title=element_text(size=14,face="bold")) +
  labs(x='Household Income in Thousands',
       y='TFR') +
  scale_color_manual(values=c('red','blue'), guide=FALSE) +
  geom_text(x=7.5,y=2.7,size=7,label='Non-White', color='black') +
  geom_text(x=8.5,y=1.7,size=7,label='White', color='black') +
  guides(fill=FALSE)
  
```


# Conclusion

We have examined our framework's errors using high-quality age pyramids and fertility data. With a few minor caveats, we conclude that indirect estimators yield accurate $TFR$ estimates when age-sex input data is accurate. 

Researchers should use caution, however, when applying our framework in data environments with potentially deficient inputs. The methods outlined above produce good estimates of $TFR$, but they require accurate age-sex counts. Underenumeration of children aged 0-4 is a well-known problem with census data [@ewbank1981age; @o2014historical; @o2015coverage], and misreporting of children aged 0-4 will yield inaccurate estimates of total fertility in our framework. Future extensions could adjust for underenumeration, either by adding an undercount multiplier to Equation (\ref{eq:s0TFR}), or by including a prior distribution for undercount in the $bTFR$ model. 

Demographers should use expert judgement when choosing which variant to employ. For instance, in populations with known or suspected high child mortality one should likely estimate $TFR$ using $iTFR^+$, $xTFR^+$, or $bTFR$ to correct for child mortality (if $q_5$ estimates are available). In choosing the three example extensions in \autoref{sec:extensions}, we chose the variant to fit both the data available and known demographic characteristics of the population. We know Africa tends to have high child mortality but we are also cautious about the accuracy of WorldPop's age-structure data. This led us to $iTFR^+$, which corrects for child mortality but does not use the detailed age distribution for women of childbearing ages. Similarly, the HMD contains high quality data but the further back in time one goes the more important uncertainty about child mortality becomes. In this situation $bTFR$ is the appropriate method. To estimate U.S. fertility conditional on household income and race, we do not have child mortality available for these populations. However, U.S. child mortality is quite low, and for small subpopulations the age structure among reproductive-age women is quite variable. These considerations make $xTFR$ the most appropriate estimator. Scholars should use similar judgements when using our framework to answer their own research questions.


Our framework's flexible data requirements allow researchers to select the variant that conforms to the available data, and offers high confidence for estimates. As we show in our example extensions, this approach opens the door to fertility analyses for many populations of interest to sociologists [@lesthaeghe14], economists [@hotz1997economics], anthropologists [@greenhalgh1995situating], epidemiologists [@paulson2002pregnancy], historians [@woods2000demography], and population geographers [@sorichetta2015high]. 

There is now increased demand for high-resolution gridded population datasets for climate change research and SDGs [@jones15;@cincotta00;@golding2017mapping]. Because our methods work well for small populations, scientists could use them to estimate small-area fertility levels and changes, as inputs to gridded population projections or for gridded fertility level datasets.

Because they rely on basic, commonly collected census data the parameter-free, scale-, time-, and species-robust techniques can estimate total fertility even in areas without detailed demographic data. We anticipate this estimation framework will open new lines of inquiry into human fertility patterns. 


# Replication Files
All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available in a replication repository located at (https://github.com/mathewhauer/iTFR-replication). The analyses were performed in *R* [@rcite]. For the $bTFR$ variant we use the *rstan* package [@rstanpackage] to inferface with Stan [@stanjss2017].

\newpage
# Supplementary material {-}

\beginsupplement


```{r make-table-hfd-countries, echo=FALSE, warning=FALSE, message = FALSE, results='asis'}
# options(knitr.kable.NA = '')
hfdcountries <- tribble(
  ~Code, ~Country,
  "AUT", "Austria",
  "BGR", "Bulgaria",
  "BLR", "Belarus",
  "CAN", "Canada",
  "CHL", "Chile",
  "HRV", "Croatia",
  "CZE", "Czechia",
  "DEN", "Denmark",
  "EST", "Estonia",
  "FIN", "Finland",
  "FRATNP", "France",
  "DEUTNP", "Germany",
  "HUN", "Hungary",
  "ISL", "Iceland",
  "ITA", "Italy",
  "JPN", "Japan",
  "LTU", "Lithuania",
  "NLD", "Netherlands",
  "NOR", "Norway",
  "POL", "Poland",
  "PRT", "Portugal",
  "RUS", "Russia",
  "SVK", "Slovakia",
  "SVN", "Slovenia",
  "ESP", "Spain",
  "SWE", "Sweden",
  "CHE", "Switzerland",
  "TWN", "Taiwan",
  "UKR", "Ukraine",
  "GBRTENW", "England and Wales",
  "GBR_NIR", "Northern Ireland",
  "GBR_SCO", "Scotland",
  "USA", "USA"
)


hmdtab <- HFD_tfr  %>%
  group_by(Code) %>%
  dplyr::summarise(Start = min(Year),
                   End = max(Year),
                   yearlength = n()) %>%
  left_join(., hfdcountries)
hmdtab2 <- hmdtab %>%
  mutate(`Data Avail. (# of Years)` = paste0(Start, "-", End, " (", yearlength,")")) %>%
  dplyr::select(Country,`Data Avail. (# of Years)` )

hmdtab2[nrow(hmdtab2)+1,] <- ""

dd2 <- cbind(hmdtab2[1:16, ], hmdtab2[17:32, ])

kable(dd2, "latex", booktabs = T, caption = "\\label{HFDCountries}\\textbf{Human Fertility Database Countries and years of data availability.}") %>%
  kable_styling(latex_options = "HOLD_position")
  
```

\newpage
